---
title: "Banana Boat STA 199 Project"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   **Identify the source of the data.** This data set comes from the CORGIS Dataset Project, who got this data from the US Energy Information Administration, a part of the US Government.

-   **State when and how it was originally collected (by the original data curator, not necessarily how you found the data).** This data has been collected by the US Energy Information Administration from 1960 to 2019 based on the consumption and expenditures of different sources of energy.

-   **Write a brief description of the observations.** The observations are of different states in different years, and have variables for many energy sources, including coal, kerosene, geothermal, etc. There are also different variables for production, consumption, expenditure, and price of these types of energy. For some of these, information was not available, in which case a 0 was reported.

-   **Address ethical concerns about the data, if any.\
    **There should be no ethical concerns about the data since the data is not specific to individuals and reports many types of used energy in the US.

## Research question

-   **A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)** Which states use the most and spend the most amount of money on more renewable energy sources, and how does this vary with geographic and cultural trends?

This question is important because as climate change is progressing at a rapid rate, it is vital to move towards more renewable energy sources in order to save our earth.

-   **A description of the research topic along with a concise statement of your hypotheses on this topic.\
    **This is important on every scale, so to look at states is a good way to see how our country is progressing towards the goal of slowing global warming. Our hypothesis is that states with large metropolitan areas, like New York, California, Florida, Massachusetts, and Pennsylvania will have larger overall expenditures, but will also spend more on renewable energy sources.

-   **Identify the types of variables in your research question. Categorical? Quantitative?\
    **The data set contained both categorical and quantitative variables. The categorical variables include "State". The remaining columns, the focus on different uses of energy, are all quantitative.

## Literature

<<<<<<< HEAD
-   **Find one published credible article on the topic you are interested in researching.** Ratner, M., Bracmort, K., Brown, P., Campbell, R. J., Holt, M., Humphries, M., Pirog, R., & Yacobucci, B. D. (2018, November 5). U.S. Energy in the 21st Century: A Primer. The Congressional Research Service. Retrieved March 10, 2023, from <https://crsreports.congress.gov/product/pdf/R/R46723>
=======
-   **Find one published credible article on the topic you are interested in researching.** Ratner, M., Bracmort, K., Brown, P., Campbell, R. J., Holt, M., Humphries, M., Pirog, R., & Yacobucci, B. D. (2018, November 5). U.S. Energy in the 21st Century: A Primer. The Congressional Research Service. Retrieved March 10, 2023, from (https://crsreports.congress.gov/product/pdf/R/R467230)
>>>>>>> 052dfcd1e3c5610b8e2533a0251a33b36d8a08b8

-   **Provide a one paragraph summary about the article.** This article dives into energy usage in the United States and goes into detail about recent trends of energy usage. The text claims that, despite energy usage remaining relatively constant, the U.S. has shifted away from non-renewable energy sources (oil and coal) and towards more renewable sources of energy. The article makes the claim that natural gas and renewable energy consumption has increased, oil and nuclear power consumption didn't change and coal decreased. In the 21st century, renewable energy consumption has almost doubled, making its increasing importance obvious. Despite this, renewable energy still only makes up 11% of U.S. energy consumption.

-   **In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.** One big difference is that the article talks about general trends in the U.S. and does discuss any of the costs. Also, the article does not take into account the factors that could provide reasoning for differences in renewable energy usage by state, which we plan to look into.

## Glimpse of data

```{r}
#| label: load-data-1

energy_states <- read.csv("data/energy.csv")
glimpse(energy_states)
```

# Data 2

## Introduction and data

-   **Identify the source of the data.** Our group found the "Graduates" data set from the CORGIS - Collection of Really Great, Interesting, Situated - Dataset Project by Austin Cory Bart of Virginia Tech University.

-   **State when and how it was originally collected (by the original data curator, not necessarily how you found the data).** This data set was originally conducted by the National Survey of Recent College Graduates from 1973 through 2010, providing demographic and career information about individuals holding a bachelor's or master's degree in science, engineering, or health field from an academic institution in America. This study was a cross-sectional biennial survey that sampled United States residents who received their degree in the prior 2 or 3 academic years and were under age 76.

-   **Write a brief description of the observations.** Each observation in this data set represents each college graduate who satisfies the aforementioned conditions, with information about their age, citizenship status, educational history, as well as other variables related to their work, family, and salary.

-   **Address ethical concerns about the data, if any.** First, because the information about each survey participant is personal, we have to keep them confidential and solely for the purpose of this study. Specifically, sensitive data such as salary and place of birth should not be shared publicly. Second, although we, as researchers, are working to find meaningful correlations purely from data, it is important to keep in mind the humanity aspect of this project, as the data represents people who come from multicultural backgrounds. Last but not least, researchers should be aware of the potential bias given by the data set, as it does not necessarily reflect what is truly going on in the workplace because of different factors such as non-response from survey takers.

## Research question

<<<<<<< HEAD
-   **A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)\
=======
-   **A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)
>>>>>>> 052dfcd1e3c5610b8e2533a0251a33b36d8a08b8
    **With an interest in how policy, jobs, and different aspects of society change during times of hardships, our group wants to analyze the effects of the Great Recession from late 2007 to 2009 on the job market in the United States. In particular, we are interested to see which majors and degree types had the highest employment rate, the change in salaries level of each major during the recession, which ethnicity groups were able to stay in the workplace and which were laid off, and the percentage of people who had to work outside of their majors during college because of the recession. All in all, these mini questions serve to answer our research interest: "How did the Great Recession affect the labor market for United States college graduates from 1973 through 2010?" Listed below are the embedded questions within this research project:

- Within U.S. college graduates from 1973 to 2010, which majors and degree types had the highest employment rate during the Great Recession?

  - How much did the salaries level of U.S. college graduates from 1973 to 2010 change during the Great Recession?

  - Which ethnic groups of U.S. college graduates from 1973 to 2010 had the highest laid-off percentage during the Great Recession?

    -   What was the percentage of U.S. college graduates who worked outside of their majors or degrees during the Great Recession?

<<<<<<< HEAD
-   **A description of the research topic along with a concise statement of your hypotheses on this topic.** Humanity has survived and grown more resilience through many challenges in history - the Spanish Flu, the Great Depression, the World Wars - and the Great Recession of 2008 is not an exception. Analyzing how people reacted to these challenges with resilience and grit will inform historians, policy makers, and anthropologists with human tendency and societal structures. Furthermore, this kind of study will give us the lessons of history to not repeat these same mistakes in the future. College students can also learn about different trends in the economy that will advise them about the degrees and majors to pursue in their post-secondary journey. On a different note, research questions on the effects of one's ethnicity in the job market will reveal discriminatory patterns in the United States against marginalized groups.\
    \
    Our team will base our studies on variables related to majors, salaries, demographics, and employment status across the time frame of 2007 to 2009. In particular, the variables and their types we will be using are:
=======
-   **A description of the research topic along with a concise statement of your hypotheses on this topic.** Humanity has survived and grown more resilience through many challenges in history - the Spanish Flu, the Great Depression, the World Wars - and the Great Recession of 2008 is not an exception. Analyzing how people reacted to these challenges with resilience and grit will inform historians, policy makers, and anthropologists with human tendency and societal structures. Furthermore, this kind of study will give us the lessons of history to not repeat these same mistakes in the future. College students can also learn about different trends in the economy that will advise them about the degrees and majors to pursue in their post-secondary journey. On a different note, research questions on the effects of one's ethnicity in the job market will reveal discriminatory patterns in the United States against marginalized groups.
  
Our team will base our studies on variables related to majors, salaries, demographics, and employment status across the time frame of 2007 to 2009. In particular, the variables and their types we will be using are:
>>>>>>> 052dfcd1e3c5610b8e2533a0251a33b36d8a08b8

- `Education.Major`: String

- `Salaries.Median`: Float

- `Demographics.Ethnicity.Asians`, `Demographics.Ethnicity.Minorities`, `Demographics.Ethnicity.Whites`, `Demographics.Gender.Male`, `Demographics.Gender.Female`: Integer

-   `Employment.Reason Working Outside Field`: Integer

Our hope is to derive meaningful relationships within these variables depending on the aforementioned research questions. Some of my hypotheses include:

- Science-related majors will have the highest employment rate during this time in the country.

- The overall salaries level of U.S. workers fell dramatically as the economy entered the Great Recession.

- Minorities groups such as Asians, Hispanics, African Americans, and females will be laid off more than other groups.

- More college graduates would have to pivot and enter an industry that had more demand than the one they majored in.

-   **Identify the types of variables in your research question. Categorical? Quantitative?
    **The majority of variables in this data set and the research questions are quantitative, which include the number of degrees, the amount of people unemployed, graduates working in academic research, and so on. The only categorical variable is the major of each college graduate.

## Literature

<<<<<<< HEAD
-   **Find one published credible article on the topic you are interested in researching.** Hoynes, H. W., Miller, D. L., & Schaller, J. (2012, March 29). Who suffers during recessions? NBER. Retrieved March 9, 2023, from <https://www.nber.org/papers/w17951>
=======
-   **Find one published credible article on the topic you are interested in researching.** Hoynes, H. W., Miller, D. L., & Schaller, J. (2012, March 29). Who suffers during recessions? NBER. Retrieved March 9, 2023, from (https://www.nber.org/papers/w17951)
>>>>>>> 052dfcd1e3c5610b8e2533a0251a33b36d8a08b8

-   **Provide a one paragraph summary about the article.** The authors of this article found that the impacts of the Great Recession caused by the crash in the real estate market from 2007 to 2009 have been greater for men, for black and Hispanic workers, for young workers, and for less educated workers than for others in the labor market. The researchers identified a common theme among recessions in the past 30 years, indicating that even though dramatic changes, including the increase of women in the labor force, were happening, the pattern of unemployment remained stable. A reason for this phenomenon is because men are more often employed in highly cyclical industries. The researchers concluded that the groups who lost in the recessions of the 1980s are the same people who experience weaker labor market outcomes even in good times. Thus, the labor market effects of the Great Recession were different from those of business cycles over the three previous decades in size and length, but not in type.

-   **In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.\
    **Our research questions delve deeper into the matter of which groups are being affected the most by recessions as we study different ethnic groups and graduates with various majors and degree types, which add on to the aforementioned research on the effects of recessions on men and women. Our research will help further inform the inequalities and unemployment patterns in the labor market in the United States and incentivize policymakers to enact meaningful changes to help the most harmed demographics during recessions.

## Glimpse of data

```{r}
#| label: load-data-2

college_graduates <- read.csv("data/graduates.csv")
glimpse(college_graduates)
```

# Data 3

## Introduction and data

-   **Identify the source of the data.** The data is provided by the Los Angeles Police Department as part of their publicly available data sets.

-   **State when and how it was originally collected (by the original data curator, not necessarily how you found the data).** It was collected starting in 2020 and includes crime up until March 1, 2023 (we are only using February 1 to May 1 of 2020). The LAPD collected the data from the original crime reports which were typed up on paper and then transcribed into the data set.

-   **Write a brief description of the observations.** The data set has 28 observations, each of which concerns information about the crime that was committed. They include what the charge of the crime was, the general area it occurred in, the date and time when it happened, as well as information about the individual who committed it. Overall, the observations are meant to paint a detailed picture of each crime and include the necessary facts to best understand what happened.

-   **Address ethical concerns about the data, if any.
    **When working with data as sensitive as crime, it is important to remember that these numbers and observations represent real crimes that occurred which had impacts on real people. The ethical concern here is to balance a high level of analysis of the data with also remembering that these are not simply numbers on a page, but rather real people who were affected. The data should be taken seriously, and a level of care with the data should be taken that remains respectful.

## Research question

<<<<<<< HEAD
-   **A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)** For our research question, we want to look at how the pandemic impacted crime in the Los Angeles area in a number of areas, including crime frequency, the severity of the crime committed/type of crime, and what the demographics committing the crime were (age, sex, race).  That is all to answer the question: How did the COVID-19 pandemic affect the crime situation in Los Angeles, particularly regarding crime frequency, severity, type, and demographics?
=======
-   **A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)** For our research question, we want to look at how the pandemic impacted crime in the Los Angeles area in a number of areas, including crime frequency, the severity of the crime committed/type of crime, and what the demographics committing the crime were (age, sex, race). That is all to answer the question: How did the COVID-19 pandemic affect the crime situation in Los Angeles, particularly regarding crime frequency, severity, type, and demographics?
>>>>>>> 052dfcd1e3c5610b8e2533a0251a33b36d8a08b8

-   **A description of the research topic along with a concise statement of your hypotheses on this topic.** This is an important question, especially now, because as the pandemic has lingered in our life for years now, we have been slowly transitioning out of the pandemic lifestyle. More and more COVID protocols are being stripped away, and society is returning to more normalcy every week, so as we make the switch, it will be interesting to see how the onset of the pandemic affected crime. It has been so long since we lived in a fully non-pandemic world, so investigating how it affected society's behavior, especially in a way as important and extreme as crime, holds a lot of significance in how we adjusted into the current world we know, and how we will react coming out of it. The larger topic here is human behavior and reaction, which we are looking at through crime statistics before and after the onset of the COVID pandemic. We want to look at key factors such as how crime tendencies may have shifted, as well as overall if there was a shift in the amount of crime committed. My hypothesis is that the start of the pandemic likely decreased crime, as well as greatly altered the types of crimes being committed. With businesses shut down, and people much more likely to simply stay home, I believe it was harder to commit crimes. Additionally, the type of crime committed likely changed to be more geared towards the at-home protocols, meaning that there was likely significantly less interpersonal crime amongst strangers.

-   **Identify the types of variables in your research question. Categorical? Quantitative?** In our data set, we are looking at both categorical as well as quantitative variables. Quantitatively, we want to look at the comparison of pre-pandemic versus pandemic crime numbers. Categorically, we also want to go more in-depth and see what the profile of the crimes being committed was.

## Literature

<<<<<<< HEAD
-   **Find one published credible article on the topic you are interested in researching.** Hill, J., Raber, G. & Gulledge, L. Down with the sickness? Los Angeles burglary and COVID-19 restrictions. J Exp Criminol (2022). <https://doi.org/10.1007/s11292-022-09522-0>
=======
-   **Find one published credible article on the topic you are interested in researching.** Hill, J., Raber, G. & Gulledge, L. Down with the sickness? Los Angeles burglary and COVID-19 restrictions. J Exp Criminol (2022). (https://doi.org/10.1007/s11292-022-09522-0)
>>>>>>> 052dfcd1e3c5610b8e2533a0251a33b36d8a08b8

-   **Provide a one paragraph summary about the article.**

The objective of this article was to analyze burglary in Los Angeles based on COVID-19 restrictions. The study found that COVID-19 restrictions significantly increased burglary in industrial areas, but decreased burglary in residential areas. The study attributed the increase in crime to the decrease in capable guardianship as COVID-19 restrictions were placed. Capable guardianship is anything that discourages crime from taking place such as alarm systems or security in stores. Therefore, it is clear that as capable guardianship decreases, crime increases. However, the study did not give an explanation on why capable guardianship decreased with COVID-19 restrictions. The study also supports routine activity theory, which states that crime occurs when there is a motivated offender, a suitable target, and a lack of capable guardianship. Thus, when white collar professionals transitioned to working at home, there was a lack of suitable targets in residential areas. However, low income and essential workers continued to work in industrial areas, providing suitable targets for crime, which is why there was more crime in industrial areas compared to residential areas.

-   **In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.** Our research question is different from this study because it analyzes all different crimes during a certain period, which is not restricted to just burglary. In addition, we will look at many other factors including sex, severity of the crime, race, age, etc.

## Glimpse of data

```{r}
#| label: load-data-3

crime_2020 <- read.csv("data/Crime_Data_from_2020_to_Present.csv")
glimpse(crime_2020)
```
